Reproducible development environments with rix

### 2023-12-14 paper.qmd
first draft of intro

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to list all the dependencies of the analysis
  clearly and ideally provide an easy way to install said dependencies.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker].{proglang}, a containerization solution. This paper will present
  the Nix package manager, and an [R].{proglang} package called [rix]{.pkg} that lowers
  Nix's learning curve for users of the [R].{proglang} programming language.

@@

Matrix products: default
BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

@@

```
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
```

The `.Rprofile` file is a file that gets read by R when starting a new session, which
means that every call to the `install.packages()` function will now install the packages
from this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be to not only
install the right packages used for the analysis, but also the right version of
[R]{.proglang}. Of course, it would be possible to install the right version manually,
but here too, there are tools that simplify the process such as @rlib2023.

There are exceptions of course, a great example of a paper that provides everything
needed to reproduce its results is @mcdermott2021. The author of this paper set up
an accompagnying Github repository to the paper^[https://github.com/grantmcdermott/skeptic-priors]
containing all the instructions to install the required software and then run the
analysis. If we take a closer look at this repository, we will notice that several
tools were used to capture the compatutational environment and make it available to
other researchers:

- The version of [R]{.proglang} was stated;
- Packages and their versions were listed and saved into an `renv.lock` file;
- A `Makefile` was used to run the whole analysis and compile the paper;
- A `Dockerfile` was used to provide the complete computational environment, including the right version of [R]{.proglang} and run the whole analysis easily.

>>>>

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to list all the dependencies of the analysis
  clearly and ideally provide an easy way to install said dependencies.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will present
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that lowers
  Nix's learning curve for users of the [R]{.proglang} programming language.

@@

Matrix products: default
BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp[...]

@@

```
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
                               __linux__/jammy/2023-06-30"))
```

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process such
as [rig]{.proglang} by the R infrastructure team [-@rlib2023]. One could thus
install the right version of [R]{.proglang} using [rig]{.proglang} allows to
easily install different versions of [R]{.proglang}, one could use to install
the right version to reproduce an analysis, and then use one of the listed
packages above to install the right library of of [R]{.proglang} packages.
However, this involves many manual steps and is thus error prone. It is also
time-consuming to do so, especially if one wants to reproduce many studies.

The final step towards the gold standard of reproducibility is to package the
right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. From this image, containers can be run to reproduce
the data product with its exact dependencies. A statistical analysis, from the
simplest to the most complex one, can be seen as such a data product that comes
with many software dependencies. *Dockerizing* an analysis consists in first
building an image: in the build step, the dependencies of the analysis have to
be installed and this can be achieved using the tools mentioned above. The
scripts to run the analysis and the data are also added to the image at build
time. Instances of that image, called containers, can then be executed, which
give access to the exact development environment originally used for the study.
It is also possible to script the building of the analysis itself, such that in
one single call to the `docker run` command, a complete analysis is reproduced.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.proglang} to the [R]{.proglang} community of users. These images
can be used as bases to build other images containing statistical analyses
more easily than starting from a bare-bones image.

An optional step towards the gold standard is to use a build automation tool
such as [Make]{.proglang} to run the whole analysis when the container is
executed. Build automation tools make it easier to run arbitrary code in a
series of well-defined steps. [R]{.proglang} programmers can use the
[targets]{.pkg} by @landau2021 as a build automation tool.

@mcdermott2021 is an example of a scientific study that reached the gold
standard of reproducibility. The author of this paper set up an accompagnying
Github repository to the
paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
instructions to install the required software and then run the analysis. If we
take a closer look at this repository, we will notice that many of the tools
previously mentioned were used to capture the compatutational environment and
make it available to other researchers:

- Packages and their versions were listed and saved into an `renv.lock` file;
- A `Makefile` was used to run the whole analysis and compile the paper;
- A `Dockerfile` was used to provide the complete computational environment, including the right version of [R]{.proglang} and run the whole analysis easily.
